---
title: "Data Manipulation"
author: "Derek Anderson"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
  html_document:
    toc: true
    toc_float: true
---
# Global Setup
```{r setup, include=TRUE}
library(tidyverse)
library(knitr)
library(markdown)
# set the default global option for displaying code chunks
knitr::opts_chunk$set(echo = TRUE)
```
# Overview

This is a document that covers the material's in [Coddy's R Data Manipulation course](https://coddy.tech/courses/data_manipulation_in_r/). Feel free to use this to help complete the certificate. 

# Modules

## Data Basics

### Understanding Data Structures

#### Vectors

```{r Vectors}
# Creating a numeric vector
numbers <- c(1, 2, 3, 4, 5)

# Creating a character vector
fruits <- c("apple", "banana", "cherry")

# Accessing elements
print(numbers[2])  # Output: 2
print(fruits[1:2])  # Output: "apple" "banana"
```

#### Matrices

```{r Matrices}
# Creating a matrix
mat <- matrix(1:6, nrow = 2, ncol = 3)

# Accessing elements
print(mat[1, 2])  # Returns the element in the 1st row and 2nd column (Output: 3)
print(mat[, 1])   # Returns the 1st column as a vector (Output: 1 2)
```

#### Data Frames

```{r}

# Creating a data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  city = c("New York", "London", "Paris")
)

# Accessing elements
print(df$name)        # Returns the column named "name" (Output: "Alice" "Bob" "Charlie")
print(df[2, "age"])   # Output: 30
```

#### Challenge & Solution

```{r Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con, n = 3))

# Process input data
data_list <- strsplit(input_data, ",")
names <- sapply(data_list, function(x) x[1])
ages <- as.numeric(sapply(data_list, function(x) x[2]))
heights <- as.numeric(sapply(data_list, function(x) x[3]))

# Create the initial data frame
df <- data.frame(
  name = names,
  age = ages,
  height = heights
)

# TODO: Write your code below
# 1. Add a new column 'height_m'
df$height_m <- heights/100
# 2. Calculate and print the average age
avg_age <- mean(df$age)
# 3. Print the name of the tallest person (in meters)
tallest_person <- df$name[df$height == max(df$height)]

# Placeholder for output
cat("Average age:", avg_age, "\n")
cat("Tallest person:", tallest_person)
```

### Dataframe Indexing

#### Example Data Frame

```{r Example Data Frame}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve"),
  age = c(25, 30, 35, 28, 22),
  city = c("New York", "London", "Paris", "Tokyo", "Sydney"),
  salary = c(50000, 60000, 75000, 55000, 45000)
)
```

#### Indexing by Position

```{r}
# Single element
df[2, 3]  # Returns the element in the 2nd row and 3rd column (character: "London")
# Entire row
df[3, ]  # Returns the 3rd row as a data frame with 1 row and all columns
# Entire column
df[, 2]  # Returns the 2nd column as a vector
df[[2]]  # Also returns the 2nd column as a vector
# Multiple rows or columns:
df[1:3, ]  # Returns the first 3 rows
df[, c(1, 3)]  # Returns the 1st and 3rd columns
```

#### Indexing by Name

```{r}
# Single column
df$age  # Returns the "age" column as a vector
df[["age"]]  # Also returns the "age" column as a vector
# Multiple columns
df[c("name", "city")]  # Returns a data frame with "name" and "city" columns
```

#### Logical Indexing

```{r}
df[df$age > 30, ]  # Returns rows where age is greater than 30
```

#### Combining Methods

```{r}
df[df$salary > 50000, "name"]  # Returns names of people with salary > 50000
```

#### Adding and Modifying Data

```{r}
# Adding a new column
df$department <- c("HR", "IT", "Finance", "Marketing", "Sales")
# Modifying existing data
df[df$name == "Alice", "salary"] <- 52000
```

#### Displaying Data Frame

```{r}
# head() previews the first rows
head(df)
# tail() previews the last rows
tail(df)
# names() shows column names
names(df)
# colnames() functions the same as names()
colnames(df)
# rownames() shows names of rows
rownames(df)
# glimpse from tidyverse, shows quick structure view
glimpse(df)
```

### Subsetting Data

Subsetting data is a crucial skill in R that allows you to extract specific portions of your data. The subset() function provides a convenient way to subset data frames.

```{r}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  city = c("New York", "London", "Paris")
)
```

#### Subsetting Rows

```{r}
# Subsetting rows
result <- subset(df, age > 28)
print(result)
```

#### Subsetting COlumns

```{r}
# Subsetting columns
result <- subset(df, select = c("name", "city"))
print(result)
```

#### Subsetting Challenge

```{r Subsetting Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con))

# Create data frame from input
df <- read.csv(text = input_data, header = TRUE, stringsAsFactors = FALSE)

# TODO: Write your code below
# 1. Subset the data frame for students older than 20
df <- subset(df, age > 20)
# 2. Select only the "name" and "score" columns
df <- subset(df, select = c("name", "score"))
# 3. Calculate the average score
# added code to round to decimals using round()
avg_score <- round(mean(df$score), 2)

# 4. Find the student with the highest score
smartest_student <- df$name[df$score == max(df$score)]

# Print results (update these with your calculations)
cat("Average score:", avg_score, "\n")
cat("Student with highest score:", smartest_student, "\n")
```

### Data Type Conversion

### The Pipe Operator in R

## Data Transformation

### Intro to dplyr

### Aggregating with dplyr

### Joining Data with dplyr
