---
title: "Data Manipulation"
author: "Derek Anderson"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
  html_document:
    toc: true
    toc_float: true
---
# Global Setup
```{r setup, include=TRUE}
library(tidyverse)
library(knitr)
library(markdown)
# set the default global option for displaying code chunks
knitr::opts_chunk$set(echo = TRUE)
```
# Overview

This is a document that covers the material's in [Coddy's R Data Manipulation course](https://coddy.tech/courses/data_manipulation_in_r/). Feel free to use this to help complete the certificate. 

# Modules

## Data Basics

### Understanding Data Structures

#### Vectors

```{r Vectors}
# Creating a numeric vector
numbers <- c(1, 2, 3, 4, 5)

# Creating a character vector
fruits <- c("apple", "banana", "cherry")

# Accessing elements
print(numbers[2])  # Output: 2
print(fruits[1:2])  # Output: "apple" "banana"
```

#### Matrices

```{r Matrices}
# Creating a matrix
mat <- matrix(1:6, nrow = 2, ncol = 3)

# Accessing elements
print(mat[1, 2])  # Returns the element in the 1st row and 2nd column (Output: 3)
print(mat[, 1])   # Returns the 1st column as a vector (Output: 1 2)
```

#### Data Frames

```{r}

# Creating a data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  city = c("New York", "London", "Paris")
)

# Accessing elements
print(df$name)        # Returns the column named "name" (Output: "Alice" "Bob" "Charlie")
print(df[2, "age"])   # Output: 30
```

#### Challenge & Solution

```{r Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con, n = 3))

# Process input data
data_list <- strsplit(input_data, ",")
names <- sapply(data_list, function(x) x[1])
ages <- as.numeric(sapply(data_list, function(x) x[2]))
heights <- as.numeric(sapply(data_list, function(x) x[3]))

# Create the initial data frame
df <- data.frame(
  name = names,
  age = ages,
  height = heights
)

# TODO: Write your code below
# 1. Add a new column 'height_m'
df$height_m <- heights/100
# 2. Calculate and print the average age
avg_age <- mean(df$age)
# 3. Print the name of the tallest person (in meters)
tallest_person <- df$name[df$height == max(df$height)]

# Placeholder for output
cat("Average age:", avg_age, "\n")
cat("Tallest person:", tallest_person)
```

### Dataframe Indexing

#### Example Data Frame

```{r Example Data Frame}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve"),
  age = c(25, 30, 35, 28, 22),
  city = c("New York", "London", "Paris", "Tokyo", "Sydney"),
  salary = c(50000, 60000, 75000, 55000, 45000)
)
```

#### Indexing by Position

```{r}
# Single element
df[2, 3]  # Returns the element in the 2nd row and 3rd column (character: "London")
# Entire row
df[3, ]  # Returns the 3rd row as a data frame with 1 row and all columns
# Entire column
df[, 2]  # Returns the 2nd column as a vector
df[[2]]  # Also returns the 2nd column as a vector
# Multiple rows or columns:
df[1:3, ]  # Returns the first 3 rows
df[, c(1, 3)]  # Returns the 1st and 3rd columns
```

#### Indexing by Name

```{r}
# Single column
df$age  # Returns the "age" column as a vector
df[["age"]]  # Also returns the "age" column as a vector
# Multiple columns
df[c("name", "city")]  # Returns a data frame with "name" and "city" columns
```

#### Logical Indexing

```{r}
df[df$age > 30, ]  # Returns rows where age is greater than 30
```

#### Combining Methods

```{r}
df[df$salary > 50000, "name"]  # Returns names of people with salary > 50000
```

#### Adding and Modifying Data

```{r}
# Adding a new column
df$department <- c("HR", "IT", "Finance", "Marketing", "Sales")
# Modifying existing data
df[df$name == "Alice", "salary"] <- 52000
```

#### Displaying Data Frame

```{r}
# head() previews the first rows
head(df)
# tail() previews the last rows
tail(df)
# names() shows column names
names(df)
# colnames() functions the same as names()
colnames(df)
# rownames() shows names of rows
rownames(df)
# glimpse from tidyverse, shows quick structure view
glimpse(df)
```

### Subsetting Data

Subsetting data is a crucial skill in R that allows you to extract specific portions of your data. The subset() function provides a convenient way to subset data frames.

```{r}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  city = c("New York", "London", "Paris")
)
```

#### Subsetting Rows

```{r}
# Subsetting rows
result <- subset(df, age > 28)
print(result)
```

#### Subsetting COlumns

```{r}
# Subsetting columns
result <- subset(df, select = c("name", "city"))
print(result)
```

#### Subsetting Challenge

```{r Subsetting Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con))

# Create data frame from input
df <- read.csv(text = input_data, header = TRUE, stringsAsFactors = FALSE)

# TODO: Write your code below
# 1. Subset the data frame for students older than 20
df <- subset(df, age > 20)
# 2. Select only the "name" and "score" columns
df <- subset(df, select = c("name", "score"))
# 3. Calculate the average score
# added code to round to decimals using round()
avg_score <- round(mean(df$score), 2)

# 4. Find the student with the highest score
smartest_student <- df$name[df$score == max(df$score)]

# Print results (update these with your calculations)
cat("Average score:", avg_score, "\n")
cat("Student with highest score:", smartest_student, "\n")
```

### Data Type Conversion

#### Checking Data Types

Before converting, it's important to know the current data type of your variable. Use the class() function to check:

```{r}
x <- 42
class(x)  # Output: "numeric"

y <- "Hello"
class(y)  # Output: "character"
```

#### Numeric to Character Conversion

Use the as.character() function to convert numeric to character:

```{r}
num <- 42
char_num <- as.character(num)
class(char_num)  # Output: "character"
print(char_num)  # Output: "42"
```

#### Character to Numeric Conversion

Use the as.numeric() function to convert character to numeric:

```{r}
char <- "3.14"
num_char <- as.numeric(char)
class(num_char)  # Output: "numeric"
print(num_char)  # Output: 3.14
```

#### Character to Factor Conversion

Use the as.factor() function to convert character to factor:

```{r}
fruits <- c("apple", "banana", "apple", "cherry")
factor_fruits <- as.factor(fruits)
class(factor_fruits)  # Output: "factor"
print(factor_fruits)  # Output: apple banana apple cherry
                      #         Levels: apple banana cherry
```

#### Factor to Chracter Conversion

Use the as.character() function to convert factor to character:

```{r}
char_fruits <- as.character(factor_fruits)
class(char_fruits)  # Output: "character"
print(char_fruits)  # Output: "apple" "banana" "apple" "cherry"
```

#### Challenge

```{r Conversion Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_string <- suppressWarnings(readLines(con))

# Convert input string to numeric vector
numeric_vector <- as.numeric(strsplit(input_string, ",")[[1]])

# TODO: Write your code below

# 1. Convert numeric_vector to a character vector
# 2. Convert the character vector to a factor
# 3. Convert the factor back to a character vector
# 4. Calculate the mean of the original numeric vector
char_vector <- as.character(numeric_vector)
factor_vector <- as.factor(char_vector)
char_from_factor <- as.character(factor_vector)
mean_value <- mean(numeric_vector)

# Print the result
print(numeric_vector)
print(char_vector)
print(factor_vector)
print(char_from_factor)
print(mean_value)
```

### The Pipe Operator in R

The pipe operator (%>%) in R enhances code readability and intuitiveness, especially when used with data manipulation packages like dplyr. It takes the output from one function and passes it as the first argument to the next, effectively meaning "and then" in the code flow.

#### Basic Usage

Without pipe:

```{r}
# result <- function2(function1(data))
# Without pipe
squared_sum <- sum(sqrt(c(1, 4, 9)))
```

With pipe:

```{r}
# result <- data %>% function1() %>% function2()
# With pipe
squared_sum <- c(1, 4, 9) %>% sqrt() %>% sum()
```

**Advantages of Using Pipes**

- Improved readability: Code flows from left to right, similar to how we read.
- Reduced need for intermediate variables.
- Easier to add or modify steps in a sequence of operations.

#### Using Pipes with Arguments

You can still specify additional arguments after the pipe:

```{r}
mtcars %>% head(n = 5)

```

In this case, mtcars is passed as the first argument to head(), and n = 5 is passed as the second argument.

Understanding the pipe operator will greatly enhance your ability to write clean, readable R code, especially when working with dplyr and other tidyverse packages.

## Data Transformation

### Intro to dplyr

The dplyr package is a powerful tool for data manipulation in R. It provides a set of functions that make it easy to transform and analyze data frames efficiently. In this lesson, we'll focus on three key functions: select(), filter(), and mutate().

#### Loading dplyr

Here is an example of how to load the package:

```{r}
# Load dplyr
library(dplyr)
```

#### select() Function

The select() function allows you to choose specific columns from a data frame:

```{r}
# Create a sample data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  city = c("New York", "London", "Paris")
)

# Select specific columns
result <- select(df, name, age)
print(result)
```
#### filter() Function

The filter() function helps you subset rows based on specific conditions:

```{r}
# Filter rows where age is greater than 28
result <- filter(df, age > 28)
print(result)
```
#### mutate() Function

The mutate() function allows you to create new columns or modify existing ones:

```{r}
# Add a new column 'age_group'
result <- mutate(df, age_group = ifelse(age < 30, "Young", "Adult"))
print(result)
```
#### arrange() Function

Ascending Order (lowest to highest): The default behavior is ascending.

```{r}
df %>%
  arrange(age)
```
Descending Order (highest to lowest): Use the desc() function to reverse the order.

```{r}
df %>%
  arrange(desc(age))
```

#### Pipe Operator %>%

dplyr introduces the pipe operator %>%, which allows you to chain multiple operations:

```{r}
result <- df %>%
  filter(age > 28) %>%
  select(name, city) %>%
  mutate(location = paste(name, "lives in", city))
print(result)
```
#### Challenge

```{r Data Transformation Challenge, eval=FALSE}
# Load required library
suppressPackageStartupMessages(library(dplyr))

# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con))

# Create data frame from input
df <- read.csv(text = input_data, stringsAsFactors = FALSE)

# TODO: Write your code below to process the data frame
process_employee_data <- function(df) {
  # Your code here
  # Use dplyr functions to:
  # 1. Select columns
  df <- select(df, name, age, salary) %>%
  # 2. Filter data
  filter(age >= 30) %>%
  # 3. Add bonus column
  mutate (bonus = ifelse(age >= 40, salary * .1, salary * .05)) %>%
  # 4. Arrange by salary
  arrange(desc(salary))
  # Replace this with your processed data frame
  return(df)  
}

# Process the data frame
result <- process_employee_data(df)

# Print the result
print(result)
```

### Aggregating with dplyr

Aggregating data is a crucial step in data analysis, allowing you to summarize and gain insights from your dataset. The dplyr package in R provides powerful functions for data aggregation, particularly group_by() and summarize(). Let's explore how to use these functions effectively.

#### group_by() Function

The group_by() function allows you to group your data based on one or more variables. This is typically the first step in data aggregation.

```{r}
# Sample data frame
df <- data.frame(
  category = c("A", "B", "A", "B", "A"),
  value = c(10, 15, 20, 25, 30)
)

# Group the data by category
grouped_df <- group_by(df, category)
print(grouped_df)
```
#### summarize() Function

After grouping, use summarize() to calculate summary statistics for each group.

```{r}
# Calculate mean value for each category
result <- summarize(grouped_df, mean_value = mean(value))
print(result)
```

#### Combining group_by() and summarize()

You can chain these functions using the pipe operator %>% for more readable code:

```{r}
result <- df %>%
  group_by(category) %>%
  summarize(mean_value = mean(value),
            sum_value = sum(value),
            count = n())
print(result)
```

#### Multiple Summary Statistics

You can calculate multiple statistics in a single summarize() call:

```{r}
result <- df %>%
  group_by(category) %>%
  summarize(
    mean_value = mean(value),
    min_value = min(value),
    max_value = max(value),
    count = n()
  )
print(result)
```

#### Aggregating Challenge

```{r Aggregating Challenge, eval=FALSE}
# Read input
con <- file("stdin", "r")
input_data <- suppressWarnings(readLines(con))

# Convert input string to data frame
suppressPackageStartupMessages(library(dplyr))

sales_data <- read.csv(text = paste(input_data, collapse = "\n"))

# TODO: Write your code below to process the sales data
# Hint: Use dplyr functions like group_by(), summarize(), arrange(), desc()

process_sales_data <- function(data) {
  # Your code here
  data <- data %>%
  group_by(category) %>%
  summarize(
    total_sales = sum(price * quantity),
    average_price = mean(price),
    items_sold = sum(quantity)
  ) %>%
  arrange(desc(total_sales))
  # Return the processed data frame
  return(data)
}

# Process the sales data
result <- process_sales_data(sales_data)

# Print the result
print(result)
```

### Joining Data with dplyr

Joining data is a crucial operation in data manipulation, allowing you to combine multiple data frames based on common columns. The dplyr package in R provides efficient functions for various types of joins. Let's explore the main join operations: inner join, left join, right join, and full join.

#### Preparing Data

First, let's create two sample data frames to demonstrate join operations:

```{r}
# Load dplyr
library(dplyr)

# Create sample data frames
employees <- data.frame(
  emp_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  department = c("HR", "IT", "Finance", "IT")
)

salaries <- data.frame(
  emp_id = c(1, 2, 3, 5),
  salary = c(50000, 60000, 55000, 65000)
)
```

#### Inner Join

An inner join returns only the rows that have matching values in both data frames:

```{r}
inner_join_result <- inner_join(employees, salaries, by = "emp_id")
print(inner_join_result)
```

#### 3. Left Join

A left join returns all rows from the left data frame and the matched rows from the right data frame:

```{r}
left_join_result <- left_join(employees, salaries, by = "emp_id")
print(left_join_result)
```

#### Right Join

A right join returns all rows from the right data frame and the matched rows from the left data frame:

```{r}
right_join_result <- right_join(employees, salaries, by = "emp_id")
print(right_join_result)
```

#### Full Join

A full join returns all rows when there is a match in either left or right data frame:

```{r}
full_join_result <- full_join(employees, salaries, by = "emp_id")
print(full_join_result)
```

#### Specifying Join Columns

If the column names are different in the two data frames, you can specify the join columns explicitly:

```{r}
# Assuming 'salaries' has 'employee_id' instead of 'emp_id'
salaries_alt <- salaries %>% rename(employee_id = emp_id)

join_result <- inner_join(employees, salaries_alt, by = c("emp_id" = "employee_id"))
print(join_result)
```

#### 7. Add New Columns

Adding or Modifying Columns with mutate() The mutate() function in dplyr allows you to add new columns or modify existing ones in a data frame. It's often used after join operations to perform calculations or transformations on the joined data.

Example:

```{r}
# Add a bonus column to the joined data
result_with_bonus <- inner_join(employees, salaries, by = "emp_id") %>%
  mutate(bonus = salary * 0.1)

# Add a performance category based on salary
result_with_category <- inner_join(employees, salaries, by = "emp_id") %>%
    mutate(performance_category = ifelse(salary < 55000, "Needs Improvement",
                                  ifelse(salary < 60000, "Meets Expectations",
                                  "Exceeds Expectations")))
# join, select(), and print result
join_result <- right_join(
  result_with_bonus,
  result_with_category %>% select(emp_id, performance_category), 
  by = "emp_id"
) # by using selct you can only add performance category to avoid duplicate name.x or name.y columns. here pipe %>% is used in the middle of the function operation
print(join_result)
```
#### Data Transformation Challenge

```{r, eval=FALSE}
# Read all input lines
con <- file("stdin", "r")
all_lines <- suppressWarnings(readLines(con))

# Find the index where the second data frame starts
separator_index <- which(grepl("^book_id,rating,user_id", all_lines))

# Split the input into two parts
books_data <- all_lines[1:(separator_index - 1)]
ratings_data <- all_lines[separator_index:length(all_lines)]

# Load required library
suppressPackageStartupMessages(library(dplyr))

# Convert string inputs to data frames
books <- read.csv(text = paste(books_data, collapse = "\n"), stringsAsFactors = FALSE)
ratings <- read.csv(text = paste(ratings_data, collapse = "\n"), stringsAsFactors = FALSE)

# TODO: Write your code below
# Perform the required join operations and calculations

# Function to perform join operations
perform_joins <- function(books, ratings) {
  # Your code here
  # Return a list containing all four join results
  list(
    inner_join = inner_join(books, ratings, by = "book_id") %>% mutate(num_ratings = ifelse(is.na(rating), 0, 1)),
    left_join = left_join(books, ratings, by = "book_id") %>% mutate(num_ratings = ifelse(is.na(rating), 0, 1)),
    right_join = right_join(books, ratings, by = "book_id") %>% mutate(num_ratings = ifelse(is.na(rating), 0, 1)),
    full_join = full_join(books, ratings, by = "book_id") %>% mutate(num_ratings = ifelse(is.na(rating), 0, 1))
  )
}
# I think this is a poor challenge because num_ratings is actually has_rating with 0 or 1 being TRUE FALSE. There were no test cases of the same book being rated twice which would actually require you to count the number of ratings.

# Call the function and store the results
join_results <- perform_joins(books, ratings)
results <- join_results

# Print the results
print(results)
```

